---
title: "Quality control of CRISPR screen counts"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quality control of CRISPR screen counts}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  dpi = 96
)
```

This vignette walks through quality control (QC) of a pooled CRISPR screen
with the `qc_*` functions in crisprFlow. It uses a small simulated screen
shipped with the package, but the same steps apply to real count data.

```{r setup}
library(crisprFlow)
```

## Inputs

The QC functions work from two tables, described in detail in
`?qc_inputs`.

**The count matrix** (`counts`) has one row per sgRNA, the columns `sgRNA`
and `Gene`, and one column of raw counts per sample. This is the layout
written by MAGeCK and guide-counter.

```{r}
counts <- exampleCounts
head(counts)
```

**The sample table** (`meta`) has one row per sample, and links each count
column (`sample_id`) to a display `label`, a biological `group` and a
`replicate`. The row order sets the order of samples in all plots, and the
levels of `group` (if it is a factor) set the order of groups. Any extra
columns, like `day` here, can be used to colour or split plots.

```{r}
meta <- exampleSampleMeta
meta
```

The example screen has a plasmid sample, and T0 (day 2), untreated and
drug-treated (day 9) samples in triplicate. `qc_check_meta()` confirms that
the sample table has the required columns and matches the count matrix:

```{r}
qc_check_meta(meta, counts)
```

**The count summary** (`summary_df`) holds the number of sequenced and
mapped reads per sample:

```{r}
summary_df <- exampleCountSummary
summary_df
```

### Reading your own data

For a real screen, the counts and summaries come from the counting
functions (e.g. `count_sgRNA_guidecounter_fastq()`), which write one
`<sample>.count.txt` and one `<sample>.countsummary.txt` file per sample.
Read and merge them with

```{r, eval = FALSE}
counts <- qc_read_count_dir("path/to/counts")
summary_df <- qc_read_summary_dir("path/to/counts")
```

The sample table is typically kept as a CSV file next to the data, e.g.

```{r, eval = FALSE}
meta <- readr::read_csv("samplesheet.csv",
                        col_types = readr::cols(replicate = "c")) |>
  dplyr::mutate(group = factor(group, levels = unique(group)))
```

## Were the counts generated against the right library?

Every metric below assumes that the counts and the library refer to the
same sgRNAs. `qc_check_library_concordance()` compares sgRNA IDs in the
counts with those of a library, by name for libraries shipped with
crisprFlow, or from a file (`library_file`). The example data contain only a
subset of the CHIP1 library, so most library sgRNAs are missing from the
counts here:

```{r}
qc_check_library_concordance(counts, qc_read_library("CHIP1"))
```

## Did reads map to the library?

The fraction of reads assigned to a library sgRNA should be consistently
high; MAGeCK-VISPR uses 65% as the minimum. Low rates point to the wrong
library, primer or stagger trimming problems, or contamination. Samples
below `min_pct_mapped` are marked "(low)":

```{r}
qc_plot_read_mapping(summary_df, meta, interactive = FALSE)
```

`Drug_3` has a low mapping rate.

Most plotting functions return an interactive plotly plot by default
(`interactive = TRUE`), which is convenient in HTML reports; this vignette
uses static plots.

## Is library representation deep and even?

`qc_sample_stats()` computes per-sample metrics from the counts: the number
of targeting sgRNAs with zero or few reads, the median count and the share
of reads on control sgRNAs.

```{r}
sample_stats <- qc_sample_stats(counts, meta, low_threshold = 30)
sample_stats
```

Any per-sample metric can be plotted with `qc_plot_sample_metric()`. The
Gini index measures how unevenly reads are spread over sgRNAs; typical
thresholds are 0.1 for early and 0.2 for late samples:

```{r}
qc_plot_sample_metric(summary_df, meta, value = "gini_index",
                      y_label = "Gini index", thresholds = c(0.1, 0.2),
                      interactive = FALSE)
```

The count distributions show the same from another angle: in each group,
the replicates' distributions should be narrow and overlap. The dotted
line marks the low-count threshold, dashed lines the medians:

```{r, fig.height = 5.5}
qc_plot_count_distribution(counts, meta, low_threshold = 30)
```

The `Drug_3` distribution is broader than those of its replicates.

## Was the library lost before selection?

A bottleneck at transduction or early in culture leaves some sgRNAs
under-represented before selection starts. `qc_guide_representation()`
calls each sgRNA detected or not at the stages you define, pooling the
samples of each stage:

```{r}
rep_data <- qc_guide_representation(
  counts,
  stages = list(Plasmid = "Plasmid", T0 = c("T0_1", "T0_2", "T0_3")),
  min_reads = 30)
qc_plot_guide_detection(rep_data)
```

Genes that lost several of their sgRNAs rest on fewer sgRNAs in the hit
analysis, so their results deserve more caution:

```{r}
qc_genes_with_lost_guides(rep_data, stage = "T0", min_frac = 0.4)
```

## Do replicates agree?

Replicates should correlate more strongly with each other than with other
groups. Correlations are computed on log2(CPM + 1) of targeting sgRNAs:

```{r, fig.height = 5.5}
qc_plot_replicate_correlation(counts, meta, interactive = FALSE)
```

`qc_replicate_correlation_table()` summarises the correlations within each
group:

```{r}
qc_replicate_correlation_table(counts, meta)
```

The PCA gives the same picture; samples should cluster by group:

```{r}
qc_plot_pca(counts, meta, interactive = FALSE)
```

Both point to `Drug_3` as an outlier, which also had a low mapping rate
and a broad count distribution: a candidate for exclusion or
re-sequencing.

## Are control sgRNAs stable?

Control sgRNAs should be phenotypically neutral, so their share of reads
should be stable across samples. A systematic shift between conditions
indicates normalization problems or selection on the controls themselves:

```{r}
qc_plot_sample_metric(sample_stats, meta, value = "pct_ctrl",
                      y_label = "% reads on control sgRNAs",
                      interactive = FALSE)
```

The heatmap shows each control sgRNA in each sample, so missing or outlier
controls stand out:

```{r, fig.height = 6}
qc_plot_control_guides(counts, meta, interactive = FALSE)
```

Controls are recognised by the `ctrl_pattern` argument (default
`"NEG_CTRL"`), which matches the control genes of all libraries shipped
with crisprFlow (e.g. `NONE_TARGETING_NEG_CTRL` and `CTRLA_NEG_CTRL`).

## Customising plots

- `colour_by`, `facet_by` and `shape_by` take any column of the sample
  table, e.g. `facet_by = "day"`.
- `colours` takes a named vector of colours for the values of the
  `colour_by` column.
- Static plots are ggplot objects and can be modified further with
  ggplot2.
- The font is set with `options(crisprFlow.font = "Arial")`.

```{r}
qc_plot_pca(counts, meta, colour_by = "group", shape_by = "replicate",
            colours = c(Plasmid = "grey40", T0 = "#0073C2",
                        Untreated = "#EFC000", Drug = "#CD534C"),
            interactive = FALSE)
```
